---
title: "Tuna gut microbiome degradation analyses"
output: html_notebook
---

# Packages 
```{r}
library(qiime2R)
library(ape)
library(Biostrings)
library(biomformat)
library(phyloseq)
library(Hmisc)
library(yaml)
library(tidyr)
library(dplyr)
library(stats)
library(utils)
library(ggplot2)
library(microbiome)
library(BiocManager)
library(DESeq2)
library(ggpubr)
library(ggsci)
library(reshape2)
library(vegan)
library(ANCOMBC)
library(limma)
library(tidyverse)
library(DECIPHER)
library(dendextend)
library(ggtext)
library(ggordiplots)
library(tibble)
library(biomeUtils)
library(rstatix)
library(emmeans)
library(multcomp)
library(MicrobiomeStat)
library(microbiomeMarker)
```

# input data from qiime2
### Experiment1
```{r}
exp1_table <- read_qza("exp1-table.qza")
exp1_metadata<-read_q2metadata("exp1-metadata.tsv")
taxonomy<-read_qza("taxonomy.qza")
```

create phyloseq object
```{r}
exp1_physeq <- qza_to_phyloseq(
    features="exp1-table.qza",
    tree="rooted-tree.qza",
    "taxonomy.qza",
    metadata = "exp1-metadata.tsv"
    )
exp1_physeq
```

```{r}
summarize_phyloseq(exp1_physeq)
```


### Experiment2
```{r}
exp2_table <- read_qza("exp2-table.qza")
exp2_metadata<-read_q2metadata("exp2-metadata.tsv")
taxonomy<-read_qza("taxonomy.qza")
```

create phyloseq object
```{r}
exp2_physeq <- qza_to_phyloseq(
    features="exp2-table.qza",
    tree="rooted-tree.qza",
    "taxonomy.qza",
    metadata = "exp2-metadata.tsv"
    )
exp2_physeq
```

```{r}
summarize_phyloseq(exp2_physeq)
```


# Data cleaning
### diversity check
```{r}
summary(sample_sums(exp1_physeq))
```

```{r}
vegan::rarecurve(t(abundances(exp1_physeq)), 
                      step = 50, label = FALSE, 
                      sample = min(rowSums(t(abundances(exp1_physeq))), 
                                   col = "blue", cex = 0.6))
```

```{r}
summary(sample_sums(exp2_physeq))
```

```{r}
vegan::rarecurve(t(abundances(exp2_physeq)), 
                      step = 50, label = FALSE, 
                      sample = min(rowSums(t(abundances(exp2_physeq))), 
                                   col = "blue", cex = 0.6))
```

### taxonomic information cleaning
Experiment1
```{r}
table(tax_table(exp1_rar)[, "Phylum"], exclude = NULL)
```

```{r}
exp1_clean <- subset_taxa(exp1_physeq, !is.na(Phylum) & !Phylum %in% c(""))
table(tax_table(exp1_clean)[, "Phylum"], exclude = NULL)
```

```{r}
table(tax_table(exp1_clean)[, "Family"], exclude = NULL)
```

```{r}
exp1_clean <- subset_taxa(exp1_clean, (Family != "mitochondria") | is.na(Family))
table(tax_table(exp1_clean)[, "Family"], exclude = NULL)
```

```{r}
print(exp1_clean)
```

Experiment2
```{r}
table(tax_table(exp2_rar)[, "Phylum"], exclude = NULL)
```

```{r}
exp2_clean <- subset_taxa(exp2_physeq, !is.na(Phylum) & !Phylum %in% c(""))
table(tax_table(exp2_clean)[, "Phylum"], exclude = NULL)
```

```{r}
table(tax_table(exp2_clean)[, "Family"], exclude = NULL)
```

```{r}
exp2_clean <- subset_taxa(exp2_clean, (Family != "mitochondria") | is.na(Family))
table(tax_table(exp2_clean)[, "Family"], exclude = NULL)
```

```{r}
print(exp2_clean)
```

# Metadata cleaning
```{r}
exp1_clean@sam_data$time <- factor(exp1_clean@sam_data$time, levels = c("0h","2h","24h","288h","576h"))
exp1_clean@sam_data$`fish-number` <- factor(exp1_clean@sam_data$`fish-number`,levels = c("1","2","3","4","5","6","7","8","9","10"))
```

```{r}
exp2_clean@sam_data$time <- factor(exp2_clean@sam_data$time, levels = c("0h","2h","24h","288h","576h"))
```

# ALpha diversity analysis - Experiment 1
```{r}
d1_clean <- meta(exp1_clean)
d1_clean$diversity_chao <- microbiome::richness(exp1_clean, "chao1")$chao1
d1_clean$eveness_sim <- microbiome::evenness(exp1_clean, "simpson")$simpson
d1_clean$low_abundance <- microbiome::rarity(exp1_clean, "low_abundance")$low_abundance
d1_clean$high_abundance <- microbiome::dominance(exp1_clean, "core_abundance")$core_abundance
```


```{r}
shapiro.test(resid(aov(diversity_chao ~ time, data = d1_clean)))
```

```{r}
summary(aov(diversity_chao~time+fish.number, data=d1_clean))
```

```{r}
TukeyHSD(aov(diversity_chao~time+fish.number, data=d1_clean), "time")
```


```{r}
shapiro.test(resid(aov(eveness_sim ~ time+fish.number, data = d1_clean))) 
```

```{r}
summary(aov(eveness_sim~time+fish.number, data=d1_clean))
```

```{r}
shapiro.test(resid(aov(low_abundance ~ time+fish.number, data = d1_clean))) 
```

```{r}
summary(aov(low_abundance~time+fish.number, data=d1_clean))
```

```{r}
TukeyHSD(aov(low_abundance~time+fish.number, data=d1_clean), "time")
```

```{r}
shapiro.test(resid(aov(high_abundance ~ time+fish.number, data = d1_clean))) 
```

```{r}
summary(aov(high_abundance~time+fish.number, data=d1_clean))
```

```{r}
TukeyHSD(aov(high_abundance~time+fish.number, data=d1_clean), "time")
```

# Alpha diversity analysis - Experiment 2
```{r}
d2_clean <- meta(exp2_clean)
d2_clean$diversity_chao <- microbiome::richness(exp2_clean, "chao1")$chao1
d2_clean$eveness_sim <- microbiome::evenness(exp2_clean, "simpson")$simpson
d2_clean$low_abundance <- microbiome::rarity(exp2_clean, "low_abundance")$low_abundance
d2_clean$high_abundance <- microbiome::dominance(exp2_clean, "core_abundance")$core_abundance
```

```{r}
shapiro.test(resid(aov(diversity_chao ~ time, data = d2_clean)))
```

```{r}
kruskal.test(d2_clean$diversity_chao, d2_clean$time)
```

```{r}
pairwise.wilcox.test(d2_clean$diversity_chao, d2_clean$time,p.adjust.method="holm")
```

```{r}
shapiro.test(resid(aov(eveness_sim ~ time+fish.number, data = d2_clean))) 
```

```{r}
kruskal.test(d2_clean$eveness_sim, d2_clean$time)
```

```{r}
pairwise.wilcox.test(d2_clean$eveness_sim, d2_clean$time,p.adjust.method="holm")
```

```{r}
shapiro.test(resid(aov(low_abundance ~ time+fish.number, data = d2_clean))) 
```

```{r}
kruskal.test(d2_clean$low_abundance, d2_clean$time)
```

```{r}
pairwise.wilcox.test(d2_clean$low_abundance, d2_clean$time,p.adjust.method="holm")
```

```{r}
shapiro.test(resid(aov(high_abundance ~ time+fish.number, data = d2_clean))) 
```

```{r}
kruskal.test(d2_clean$high_abundance, d2_clean$time)
```


# Beta diversity analysis - Experiment 1
###TSS normalization
```{r}
exp1_norm <- normalize(exp1_clean, "TSS")
```

###PCoA
PCOA
```{r}
set.seed(119)
exp1_ord <- ordinate(exp1_norm, "PCoA", "bray")
plot_ordination(exp1_norm, exp1_ord, color = "fish.number", shape = "time") +
                geom_point(size = 5) +scale_color_jco() +scale_shape_manual(values = c(4,16,15,17,18)) +labs(shape="Time", colour="Fish ID")+theme_bw()
```

```{r}
exp1_norm_otu <- abundances(exp1_norm)
exp1_norm_meta <- data.frame(sample_data(exp1_norm))
unifrac.dist.exp1 <- UniFrac(exp1_norm, 
                        weighted = TRUE, 
                        normalized = TRUE,  
                        parallel = FALSE, 
                        fast = TRUE)

adonis2(unifrac.dist.exp1 ~ time, data = exp1_norm_meta)
```

```{r}
exp1_disper <- betadisper(unifrac.dist.exp1, exp1_norm_meta$time)
permutest(exp1_disper, pairwise = TRUE)
```


# Beta diversity analysis - Experiment 2
###TSS normalization
```{r}
exp2_norm <- normalize(exp2_clean, "TSS")
```

###PCoA
PCOA
```{r}
set.seed(119)
exp2_ord <- ordinate(exp2_norm, "PCoA", "bray")
plot_ordination(exp2_norm, exp2_ord, color = "time") +
                geom_point(size = 5) +scale_color_jco() + labs(color="Time")+theme_bw()
```

```{r}
exp2_norm_otu <- abundances(exp2_norm)
exp2_norm_meta <- data.frame(sample_data(exp2_norm))
unifrac.dist.exp2 <- UniFrac(exp2_norm, 
                        weighted = TRUE, 
                        normalized = TRUE,  
                        parallel = FALSE, 
                        fast = TRUE)

adonis2(unifrac.dist.exp2 ~ time, data = exp2_norm_meta)
```

```{r}
exp2_disper <- betadisper(unifrac.dist.exp2, exp1_norm_meta$time)
permutest(exp2_disper, pairwise = TRUE)
```





